---
title: "Automated pregating using cyDefine"
author: "Søren H Dam"
date: "Last updated: `r format(Sys.Date(), '%Y.%m.%d')`"
output: 
  prettydoc::html_pretty:
  theme: hpstr
  highlight: vignette
vignette: >
  %\VignetteIndexEntry{pregating}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  out.width = "100%"
)
```

## Phenotype transfer of non-pregated cells with cyDefine

The purpose of pregating is to ensure the downstream analyses solely concern intact, living, singlet cells.

In this vignette, we demonstrate that cyDefine is capable of automating this process by labeling otherwise pregated cells as *unassigned*. Thus, cyDefine can alleviate another tedious process.

For the demonstration, we replicate a pregating strategy shown in [this](https://biosurf.org/cytof_data_scientist.html#35_Pre-gating) primer on biosurf.org using the same data.
The first step is to download the sample *081216-Mike-HIMC ctrls-001_01_normalized.fcs* from FlowRepository (ID: [FR-FCM-ZYAJ](https://flowrepository.org/id/FR-FCM-ZYAJ)).

## Dependencies

```{r packages}
library(MASS)
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
library(patchwork)
library(grid)
suppressPackageStartupMessages(library(ComplexHeatmap))
suppressPackageStartupMessages(library(circlize))
library(cyCombine, warn.conflicts = FALSE)
library(cyDefine, warn.conflicts = FALSE)
```


## Prepare data

We load the sample using functions from *cyCombine*.

```{r prepare_data}
# Define markers
markers <- c("CD57", "CD19", "CD4", "CD8", "IgD", "CD11c", "CD16", "CD3", "CD38", "CD27", "CD14", "CXCR5", "CCR7", "CD45RA", "CD20", "CD127", "CD33", "CD28", "CD161", "TCRgd", "CD123", "CD56", "HLADR", "CD25")
pregating_channels <- c("Bead", "DNA1", "DNA2", "Dead", "Event_length")

# Compile fcs file into flowSet
fs <- compile_fcs("../data/ZYAJ", pattern = "001_01_normalized.fcs")

# Create panel from flowSet parameters
panel <- tibble(channel = fs[[1]]@parameters@data$name,
                antigen = fs[[1]]@parameters@data$desc) |> 
  mutate(antigen = case_when(is.na(antigen) ~ channel,
                             TRUE ~ antigen))

# Convert to data.frame
raw <- prepare_data(
  flowset = fs, 
  panel = panel, panel_antigen = "antigen", panel_channel = "channel", 
  markers = c(markers, pregating_channels[1:4]), # Include pregating channels in the transformation
  .keep = TRUE, # Keep all pregating channels
  down_sample = FALSE,
  derand = FALSE # FALSE for plotting
  )

raw
```

## Pregating

In this pregating, we keep gated cells, but label them as either *Beads, Debris, Dead, or Doublets* for the analysis.

### Gating for cells

Gating on the DNA1 and Bead markers, we can remove beads, and some bead/cell doublets and debris.

```{r gate_cells}

# Generating a plot for gating visualization
p <- ggplot2::ggplot(raw |> dplyr::sample_n(20000), aes(x = Bead, y = DNA1)) +
      geom_point(color = 'grey', size = 0.05, alpha = 0.6) +
      geom_density2d(h = max(bandwidth.nrd(raw$Bead), bandwidth.nrd(raw$DNA1))) +
      theme_bw()

# Set gate boundaries
left <- 0; right <- 2; lower <- 5; upper <- 8.5

# Gating cells
cells <- raw[raw$Bead < right,]; cells <- cells[cells$DNA1 > lower & cells$DNA1 < upper,]

# Add gate to plot
p + geom_rect(xmin = left, xmax = right, ymin = lower, ymax = upper, fill = NA, linetype = 1, color = 'red') +
  annotate('text', x = right + 1, y = mean(c(lower, upper)), label = paste("Cells: ", round((nrow(cells)/nrow(raw))*100, digits=2), "%", sep=""), color = 'red')

# Label in dataframe
raw <- raw |> 
  mutate(
    pregating = case_when(
      Bead > right ~ "Beads",
      DNA1 > upper ~ "Doublets",
      DNA1 < lower ~ "Debris",
      TRUE ~"Cells"
))
```

### Gating for intact cells

Gating on DNA1 and DNA2 helps identify intact cells.

```{r gate_intact}
# Gene rating a plot for gating visualization
p <- ggplot2::ggplot(cells |> dplyr::sample_n(10000), aes(x = DNA1, y = DNA2)) +
      geom_point(color = 'grey', size = 0.05, alpha = 0.6) +
      coord_cartesian(xlim = c(5,8), ylim = c(5.5,8.5)) +
      geom_density2d() +
      theme_bw()

# Set gate boundaries
left <- 5.75; right <- 7.2; lower <- 6.35; upper <- 7.8

# Gating intact cells
intact <- cells[cells$DNA1 > left & cells$DNA1 < right,]; intact <- intact[intact$DNA2 > lower & intact$DNA2 < upper,]

# Add gate to plot
p + geom_rect(xmin = left, xmax = right, ymin = lower, ymax = upper, fill = NA, linetype = 1, color = 'red') +
  annotate('text', x = right + 0.4, y = mean(c(lower, upper)), label = paste("Intact cells: ", round((nrow(intact)/nrow(cells))*100, digits=2), "%", sep=""), color = 'red')

# Label in dataframe
raw <- raw |> 
  mutate(
    pregating = case_when(
      pregating != "Cells" ~ pregating,
      DNA1 > right | DNA2 > upper ~ "Doublets",
      DNA1 < left | DNA2 < lower ~ "Debris",
      TRUE ~ "Cells"
))

```

### Gating for singlets

Event length helps remove doublets.

```{r gate_singlets}
# Generating a plot for gating visualization
p <- ggplot2::ggplot(intact |> dplyr::sample_n(10000), aes(x = Eventlength, y = DNA1)) +
      geom_point(color = 'grey', size = 0.05, alpha = 0.6) +
      coord_cartesian(ylim = c(0,8)) +
      geom_density2d() +
      theme_bw()

# Set gate boundaries
left <- 12; lower <- 5.65; right <- 22; upper <- 7.35

# Gating intact singlets
singlets <- intact[intact$Eventlength > left & intact$Eventlength < right,]

# Add gate to plot
p + geom_rect(xmin = left, xmax = right, ymin = lower, ymax = upper, fill = NA, linetype = 1, color = 'red') +
  annotate('text', x = mean(c(left, right)), y = lower - 0.5, label = paste("Intact singlets: ", round((nrow(singlets)/nrow(intact))*100, digits=2), "%", sep=""), color = 'red')

# Label in dataframe
raw <- raw |> 
  mutate(
    pregating = case_when(
      pregating != "Cells" ~ pregating,
      Eventlength >= right ~ "Doublets",
      Eventlength <= left ~ "Debris",
      TRUE ~ pregating
))


```


### Gating for live cells

Finally, we identify deads cells using the viability stain.

```{r gate_live}
# Generating a plot for gating visualization
p <- ggplot2::ggplot(singlets |> dplyr::sample_n(20000), aes(x = Dead, y = DNA1)) +
      geom_point(color = 'grey', size = 0.05, alpha = 0.6) +
      coord_cartesian(ylim = c(0,8)) +
      geom_density2d() +
      theme_bw()

# Set gate boundaries
left <- 0; lower <- 5.65; right <- 3; upper <- 7.35

# Gating live intact singlets
live <- singlets[singlets$Dead < right,]

# Add gate to plot
p + geom_rect(xmin = left, xmax = right, ymin = lower, ymax = upper, fill = NA, linetype = 1, color = 'red') +
  annotate('text', x = mean(c(left, right)), y = lower - 0.5, label = paste("Live intact singlets: ", round((nrow(live)/nrow(singlets))*100, digits=2), "%", sep=""), color = 'red')

# Label in dataframe
raw <- raw |> 
  mutate(
    pregating = case_when(
      pregating != "Cells" ~ pregating,
      Dead >= right ~ "Dead",
      TRUE ~ "Cells"
))
```
This pregating leaves us with the following number of kept and discarded cells:

```{r pregating_result}
table(raw$pregating)
```




## Adapt PBMC reference

With the pregating in place, we select a reference for our analysis.
Having a manually gated sample from the same experiment would be ideal. However, *cyDefine* enables us to use a universal reference.
Therefore, we use the universal PBMC reference prebuilt in *cyDefine*.

In this step, we adapt the universal PBMC reference with our sample because the panels are not identical. In the adaptation, we define the random forest parameter *mtry* as half the available markers. *mtry* determines how many markers that are randomly selected at each branching point in the model.

```{r adapt}
# Load reference
reference <- get_reference(path = "../data/")

# Map marker names to reference markers
raw_mapped <- map_marker_names(ref_markers = pbmc_markers,
  using_pbmc = TRUE, query = raw,
  query_markers = markers
)

# Redefine the marker list
non_markers <- c(non_markers, colnames(raw)[!colnames(raw) %in% markers])
markers <- get_markers(raw_mapped)

# Define RF paramters
mtry <- ceiling(length(markers)/2)

# Adapt reference
reference_adapted <- adapt_reference(
  reference = reference,
  markers = markers, 
  using_pbmc = TRUE,
  num.threads = 4,
  mtry = mtry)
```

In this case, five groups of cell populations could not be accurately separated using the smaller marker panel. Therefore, *cyDefine* merges those populations.
You can plot a diagram of the merged populations:


```{r adapt_diagram}
plot_diagram(reference_adapted)
```


## Classify all live cells

Now, we classify all live cells, i.e., cells retained after the pregating.
We leave *identify_unassigned = TRUE* to detect outliers, e.g., beads, debris, dead, and doublets.

```{r classify_live}
classified_live <- cyDefine(
  reference = reference_adapted,
  query = raw_mapped |> filter(pregating == "Cells"),
  markers = markers, 
  using_pbmc = TRUE,
  adapt_reference = FALSE,
  batch_correct = TRUE,
  norm_method = "scale",
  xdim = 6, ydim = 6,
  identify_unassigned = TRUE,
  num.threads = 4,
  mtry = mtry,
  num.trees = 500,
  train_on_unassigned = FALSE,
  seed = 332,
)


```

## Classify non-pre-gated cells

Then, we do the exact same, but without removing cells from the pregating.
We use the exact same settings to make it as comparable as possible.

```{r classify_raw}

classified_raw <- cyDefine(
  reference = reference_adapted,
  query = raw_mapped,
  markers = markers, 
  using_pbmc = TRUE,
  adapt_reference = FALSE,
  batch_correct = TRUE,
  norm_method = "scale",
  xdim = 6, ydim = 6,
  identify_unassigned = TRUE,
  num.threads = 4, 
  mtry = mtry,
  num.trees = 500,
  train_on_unassigned = FALSE,
  seed = 332,
)

```
### Exclude tiny populations

Because there is a tiny amount of ILC cells detected, we remove those cells.
Live ILC: `r sum(classified_live$query$predicted_celltype == "ILC")`. Raw ILC: `r sum(classified_raw$query$predicted_celltype == "ILC")`.

```{r rm_ilc}
classified_raw$query <- classified_raw$query |> 
  filter(predicted_celltype != "ILC")
classified_live$query <- classified_live$query |> 
  filter(predicted_celltype != "ILC")

```



## UMAP of celltype predictions

The first visualization we use is a UMAP comparing the predicted celltypes.
The UMAPs will look different because the filtered and unfiltered dataset are integrated separately with the reference, and the plots are generated using separate samplings.

```{r plot_live}
# Define celltype colors
celltype_colors <- get_distinct_colors(unique(classified_raw$query$predicted_celltype), 
                                       add_unassigned = TRUE)

umap_live <- cyDefine::plot_umap(
  classified_live$query,
  markers = markers,
  col = "predicted_celltype",
  title = "Pre-gated cells",
  colors = celltype_colors,
  down_sample = TRUE,
  sample_n = 25000
  )

```

```{r plot_raw}
umap_raw <- cyDefine::plot_umap(classified_raw$query,
          markers = markers,
          col = "predicted_celltype",
          title = "All cells",
          colors = celltype_colors,
          down_sample = TRUE,
          sample_n = 25000)

```

```{r umap,  fig.height=10,  fig.width=12}
(umap_live + umap_raw) + plot_layout(guides = "collect")
```

### UMAP of pregated cells

Here, we visually compare the pregating labels with the celltype predictions. We hope to see that most pregated cells are grouped in UMAP space, and those groups are correctly identified as "unassigned".

That is exactly what we see, especially for beads and the clusters of doublets and dead cells. Debris and many doublets appear harder to discern, at least from a UMAP.

```{r}
pregating_colors <- get_distinct_colors(unique(c(classified_raw$query$predicted_celltype, classified_raw$query$pregating)), add_unassigned = TRUE)

umap_pregating <- cyDefine::plot_umap(
  classified_raw$query,
  classified_raw$query,
  markers = markers,
  build_umap_on = "reference",
  ref_col = "predicted_celltype",
  query_col = "pregating",
  title = c("Prediction", "Pregating"),
  colors = pregating_colors,
  down_sample = TRUE,
  sample_n = 25000)
umap_pregating

```




## Celltype abundances

We also hope to see that the proportion of celltypes remain identical between the two analysis strategies. We remove unassigned cells, as the raw dataset contain many more of those.

```{r abundances, fig.height=11}
ab_live <- classified_live$query |> 
  filter(predicted_celltype != "unassigned") |>
  pull(predicted_celltype) |> 
  plot_abundance(colors = celltype_colors, title = "Cell type abundances - pregated") 

ab_raw <- classified_raw$query |> 
  filter(predicted_celltype != "unassigned") |>
  # filter(pregating == "Cells") |>
  pull(predicted_celltype) |> 
  plot_abundance(colors = celltype_colors, title = "Cell type abundances - non-pregated")

(ab_live / ab_raw) & ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, vjust = 1))
```

## Heatmap of prediction differences

Lastly, we demonstrate the accuracy of the predictions with a heatmap comparing the predictions of live and raw, where the live pregated cells are appended with labels.

```{r pheatmap, eval=FALSE,include=FALSE}
library(pheatmap)

query <- classified_raw$query |> 
  left_join(classified_live$query |> dplyr::select(id, predicted_celltype) |> rename(celltype = predicted_celltype), by = "id") |> 
  mutate(celltype = case_when(
    is.na(celltype) ~ pregating,
    TRUE ~ celltype
  ))


regular_levels <- setdiff(
  unique(query$celltype),
  c("Doublets", "Beads", "Dead", "Debris", "unassigned")
)

ordered_levels <- c("unassigned", "Beads", "Dead", "Debris", sort(regular_levels))

# Turn `celltype` into a factor with that order
query$celltype <- factor(query$celltype, levels = ordered_levels)

conf_mat <- table(query$celltype, query$predicted_celltype)
row_totals   <- rowSums(conf_mat)                # total cells per true type
pct_mat <- sweep(conf_mat, 1, row_totals, FUN = "/") * 100
pct_mat <- round(pct_mat, 1)


display_mat <- matrix(
  ifelse(t(pct_mat) == 0, "",
         paste0(round(t(pct_mat), 1), "%")),
  nrow = nrow(t(pct_mat)),
  dimnames = dimnames(t(pct_mat))
)

heatmap <- pheatmap(
  t(pct_mat),
  cluster_rows = FALSE,          # keep the order we just set
  cluster_cols = FALSE,
  display_numbers = display_mat,
  number_color = "black",
  fontsize_number = 9,
  color = colorRampPalette(c("#f7fbff", "#08519c"))(100),

  # Axis labels
  main = "Cell type prediction on non-pregated data",
  annotation_legend = FALSE,    # hide the extra legend box
  legend = TRUE,
  legend_breaks = c(0, 25, 50, 75, 100),
  legend_labels = c("25", "50", "75", "100"),
  labels_col = rownames(pct_mat),   # explicit row labels (optional)
  labels_row = colnames(pct_mat),   # explicit column labels (optional)
  border_color = "white",

  # Adding titles for the axes – pheatmap does not have dedicated
  # arguments, so we use `grid::grid.text` after the plot is drawn.
  silent = FALSE                # make sure the plot is rendered now
)
heatmap
```

```{r pheatmap2, eval=FALSE,include=FALSE}
library(pheatmap)
library(grid)

display_mat <- matrix(
  ifelse(t(pct_mat) == 0, "",
         paste0(round(t(pct_mat), 1), "%")),
  nrow = nrow(t(pct_mat)),
  dimnames = dimnames(t(pct_mat))
)

heatmap2 <- pheatmap(
  t(pct_mat),
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  display_numbers = display_mat,   # <-- use the matrix with blanks for zeros
  number_color = "black",
  fontsize_number = 7,
  color = colorRampPalette(c("#f7fbff", "#deebf7", "#c6dbef",
                              "#9ecae1", "#6baed6", "#4292c6",
                              "#2171b5", "#08519c", "#08306b"))(100),
  cellwidth  = 18,
  cellheight = 18,
  fontsize      = 9,
  fontsize_row  = 8,
  fontsize_col  = 8,
  main = "Cell type prediction on unnormalized, non-pregated data",
  legend = TRUE,
  legend_breaks = c(0, 25, 50, 75, 100),
  legend_labels = c("0", "25", "50", "75", "100"),
  angle_col = 315,
  border_color = "white",
  silent = FALSE
)

# Add axis titles using grid (pheatmap doesn't support them natively)
grid.text("Manually gated cell type",
          x = 0.55, y = 0.01,
          gp = gpar(fontsize = 10, fontface = "plain"))

grid.text("Predicted cell type",
          x = 0.01, y = 0.5,
          rot = 90,
          gp = gpar(fontsize = 10, fontface = "plain"))
```

```{r cheatmap, fig.height=15, fig.width=22, out.width = "150%"}
query <- classified_raw$query |> 
  left_join(classified_live$query |> dplyr::select(id, predicted_celltype) |> rename(celltype = predicted_celltype), by = "id") |> 
  mutate(celltype = case_when(
    is.na(celltype) ~ pregating,
    TRUE ~ celltype
  ))


regular_levels <- setdiff(
  unique(query$celltype),
  c("Doublets", "Beads", "Dead", "Debris", "unassigned")
)

ordered_levels <- c("unassigned", "Doublets", "Beads", "Dead", "Debris", sort(regular_levels))

# Turn `celltype` into a factor with that order
query$celltype <- factor(query$celltype, levels = ordered_levels)

conf_mat <- table(query$celltype, query$predicted_celltype)
row_totals   <- rowSums(conf_mat)           
pct_mat <- sweep(conf_mat, 1, row_totals, FUN = "/") * 100
pct_mat <- round(pct_mat, 1)



plot_mat <- t(pct_mat)

display_mat <- matrix(
  ifelse(plot_mat == 0, "",
         paste0(round(plot_mat, 1), "%")),
  nrow = nrow(plot_mat),
  dimnames = dimnames(plot_mat)
)

col_fun <- colorRamp2(
  c(0, 25, 50, 75, 100),
  c("#f7fbff", "#9ecae1", "#4292c6", "#2171b5", "#08519c")
)

ht <- Heatmap(
  plot_mat,
  col = col_fun,
  na_col = "#f7fbff",
  
  cell_fun = function(j, i, x, y, width, height, fill) {
    val <- plot_mat[i, j]
    if (val != 0) {
      grid.text(
        paste0(round(val, 1), "%"),
        x, y,
        gp = gpar(fontsize = 10, col = "black")
      )
    }
  },
  
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  

  row_names_side = "left",
  
  column_title = "Predicted celltypes of live cells + pregated labels",
  # column_title_side = "bottom",
  column_title_gp = gpar(fontsize = 12),
  
  row_title = "Predicted celltypes on non-pregated cells",
  row_title_side = "right",
  row_title_gp = gpar(fontsize = 12),
  
  column_names_rot = 45,
  column_names_gp = gpar(fontsize = 12),
  row_names_gp = gpar(fontsize = 12),
  
  # width  = unit(18 * ncol(plot_mat), "pt"),
  # height = unit(18 * nrow(plot_mat), "pt"),
  
  rect_gp = gpar(col = "white", lwd = 1),
  
  heatmap_legend_param = list(
    title = "Percentage of celltype\n agreement",
    at = c(25, 50, 75, 100),
    labels = c("25", "50", "75", "100"),
    color_bar = "continuous",
    legend_direction = "vertical"
  )
)

draw(ht,
     column_title = "Cell type prediction on non-pregated vs pregated data",
     column_title_gp = gpar(fontsize = 18, fontface = "bold"),
     padding = unit(c(5, 5, 5, 5), "mm"))
```

